---
title: "R Notebook"
output: html_notebook
---

```{r}
# Loads the libraries to perform functions
library(tidyverse)
set.seed(123)
```

# Step 1. Dateset

Since real data of different carriers here in Australia aren't publicy available (without breaking privacy laws) - I will create a syntethic dataset of carrier for our contract logistics firm that is highly inspired by data I have found online. Our dataset has the following columns:

- carrier_id: The identification number a carrier
- month: Month of when data was captured
- origin_state


- Cost: Price per shipment
- Performance: The on time rate in percentage
- Reliability: The overall damage rate in percentage
- Speed: The average transit days for delivery
- Scale: The fleet size, how big the company?
- Reputation: The average review score from 1 (lowest) to 5 (highest)
- Service: The tracking quality from 1 (lowest) to 5 (highest)
- Lane Distance (Operational Geography): Longer lanes are more expensive, interstate lanes may include higher delay risks
- Time: Month when data was captured
- Weather: Average weather condition

The below creates the dataset, it will have 2000 entries (rows) and 9 different attributes (columns):

```{r}
# Dataset attributes
n_carriers = 20

# Creates the simulated dataset
carrier_df <- data.frame(
  carrier_id = c(1,2,3),
  month = c(1,2,3),
  origin_state = c(1,2,3),
  destination_state = c(1,2,3),
  lane_distance_km = c(1,2,3),
  lane_type = c(1,2,3),
  price_per_shipment = c(1,2,3),
  on_time_rate = c(1,2,3),
  damage_rate = c(1,2,3),
  avg_transit_days = c(1,2,3),
  fleet_size = c(1,2,3),
  review_score = c(1,2,3)
)

carrier_df

```

```{r}
# COLUMN 1 : carrier_id : Create the dataframe and the carrier_id column
carrier <- data.frame(
  carrier_id = paste("Carrier_", sprintf("%02d", 1:n_carriers)))

carrier |> head(3) # check result
```


```{r}
# COLUMN 2 : month : Add the month variable - each carrier will be allocated to each month once
carrier <- tidyr::crossing(
  carrier_id = paste("Carrier_", sprintf("%02d", 1:n_carriers)),
  month = seq(as.Date("2020-07-01"), as.Date("2020-12-01"), by = "month")
)

carrier |> head(3) # check result
```


```{r}
# COLUMN 3 : origin_state : Add origin_state - (mainly) 80% VIC / 10% ACT / 5% SA and 5% NSW
all_origin_states = c("VIC", "SA", "ACT", "NSW")
all_origin_states_prop = c(0.75, 0.1, 0.15, 0.1)

carrier_origin <- tibble(
  carrier_id = paste("Carrier_", sprintf("%02d", 1:n_carriers)),
  origin_state = sample(all_origin_states, size = n_carriers, prob = all_origin_states_prop,
                        replace=TRUE)
)

## Join the table to our master dataframe
carrier <- carrier |> left_join(carrier_origin, by = 'carrier_id')

carrier |> dplyr::slice_sample(prop=1) |> head(3) # check result
```


```{r}
# COLUMN 4 : destination_state : Where did the carrier go to?
all_destination_states = c(all_origin_states, "QLD", "NT", "WA", "TAS")
all_destination_states_prop = c(0.13, 0.11, 0.05, 0.27, 0.22, 0.04, 0.04, 0.14)

carrier <- carrier |> dplyr::mutate(
  destination_state = sample(all_destination_states, size = nrow(carrier), 
                             prob = all_destination_states_prop, replace = TRUE))

carrier |> dplyr::slice_sample(prop=1) |> head(3) # check result
```


```{r}
# COLUMN 5 : lane_distance_km : How far awway was the delivery from Tullamarine VIC?
## Create a lookup table for all states - their base km distance from Tullamarine - and the 
## additional km that can occur in that state (E.g. Albury NSW is 320km but Moorebank NSW 820km)
distance_lookup <- tibble(
  destination_state = all_destination_states,
  base_km = c(30, 730, 470, 300, 1700, 2700, 3400, 600),
  extra_km_max = c(370, 370, 70, 900, 1000, 1200, 1000, 200)
)

## Add distance_lookup to our carrier dataset while keeping the distances in mind
carrier <- carrier |> left_join(distance_lookup, by = 'destination_state') |> 
  mutate(lane_distance_km = round(base_km + runif(n(), 0, extra_km_max), 0)) |> 
  dplyr::select( -c(base_km, extra_km_max))

carrier |> dplyr::slice_sample(prop=1) |> head(3) # check result
```


```{r}
# COLUMN 6 : Add the lane_type : If delivery is intrastate or interstate
carrier <- carrier |> mutate(
  lane_type = if_else(destination_state == "VIC", "Intrastate", "Interstate"),
  lane_type = factor(lane_type, levels = c("Intrastate", "Interstate"))
)

carrier |> head(3) # check result
```


```{r}
# Column 7 : price_per_shipment
## Let's create a carrier specific surcharge and randomise their costs (between $0.10 and $0.20/km 
## - this will have a major effect on the carrier overall costs
carrier_pricing <- tibble(
  carrier_id = unique(carrier$carrier_id),
  carrier_handling_surcharge = round(runif(length(unique(carrier$carrier_id)), 10, 100), 2),
  carrier_cost_per_km = round(runif(length(unique(carrier$carrier_id)), 0.10, 0.20), 2))

# Join the data to our main table
carrier <- carrier |> left_join(carrier_pricing, by = "carrier_id")


carrier <- carrier |> mutate(
  price_per_shipment = carrier_cost_per_km * lane_distance_km + carrier_handling_surcharge + 
    if_else(lane_type == "Interstate", 100, 0)) |> 
  select( 
    -c(carrier_handling_surcharge, carrier_cost_per_km))

# # Sanity check - uncomment to reveal the solution - Outcome: data is realistic and plausible
# ## Check minimum, maximum and average price per shipment
# carrier |> summarise(
#   min_price = min(price_per_shipment),
#   max_price = max(price_per_shipment),
#   mean_price = mean(price_per_shipment))
# 
# # What trip was the cheapest delivery in our dataset?
# carrier |> filter(price_per_shipment == min(price_per_shipment))
# 
# # What is the relationship between lane distance in km and price per shipments? Are there any unrealstic deliveries? 
# ## Outcome: NO
# ggplot(carrier, aes(x=lane_distance_km, y=price_per_shipment)) + 
#   geom_line()

carrier |> head(3) # check result
```




```{r}
# COLUMN 8 : on_time_rate : What is the percentage of on-time deliveries?
## Feature Engineer (Create Columns) that influence the calculation of on_time_rate
carrier_performance <- carrier |> 
  group_by(carrier_id) |> 
  summarise(
   mean_lane_distance = mean(lane_distance_km),
   mean_price_per_shipment = mean(price_per_shipment))

# Create new columns (Min-Max-Scaled versions of above two columns) : Will be used to influence on_time_rate
carrier_performance <- carrier_performance |> 
  mutate(
    mean_lane_distance_scaled = 
      (carrier_performance$mean_lane_distance - min(carrier_performance$mean_lane_distance)) / 
      (max(carrier_performance$mean_lane_distance) - min(carrier_performance$mean_lane_distance)),
    
    mean_price_per_shipment_scaled = 
      (carrier_performance$mean_price_per_shipment - min(carrier_performance$mean_price_per_shipment)) /
      (max(carrier_performance$mean_price_per_shipment) - min(carrier_performance$mean_price_per_shipment))
    ) |> 
  select(
    carrier_id, mean_lane_distance_scaled, mean_price_per_shipment_scaled)

carrier_performance |> head(3) # Shows the first 4 carrier with scaled distance and price
```

```{r}
# Joins those new variables to our main data - creates the new variable
carrier <- carrier |> left_join(carrier_performance, by = "carrier_id") |> mutate(
  on_time_rate =  
    0.80 + # base line percent points 80% reliable
    0.08 * (1 - mean_lane_distance_scaled) + # we weight distance twice as much as 
    0.04 * (1 - mean_price_per_shipment_scaled) + # we weight price
    rnorm(n(), mean = 0, sd = 0.02), # add some randomess to the performance/ month
  on_time_rate = round(on_time_rate*100, 2)) |> 
  select(-c(mean_price_per_shipment_scaled))

# # Sanity check if performance is realtistic - min 76%, max 96% and average 85.5%
# carrier |> group_by(carrier_id) |> 
#   summarise(
#   min = min(on_time_rate),
#   max = max(on_time_rate),
#   avg = mean(on_time_rate))
 
carrier |> dplyr::slice_sample(prop=1) |>  head(4) # inspect data
```






```{r}
# COLUMN 9 : damage_rate : Percentage of occuring damages
carrier <- carrier |> mutate(
  damage_rate = 
    0.01 + # base damage rate of 1%
    0.03 * mean_lane_distance_scaled + # distance increases risk
    rnorm(n(), mean = 0, sd = 0.005), # random operational noise
  # clamp our results to a minimum of 0.2% (for excellent carrier) and 5% (for bad ones)
  damage_rate = round(pmin(pmax(damage_rate, 0.002), 0.05) * 100, 4)) |> # multipled by 100 to represent actual percentage points
  select( -c(mean_lane_distance_scaled)) # drop the variable

# # Sanity check if values are realistic
# carrier |> group_by(carrier_id) |> 
#   summarise(
#     min = min(damage_rate), max = max(damage_rate), avg = mean(damage_rate))

carrier |> slice_sample(prop = 1) |> head(4)
```


```{r}
# COLUMN 10 : avg_transit_days
carrier <- carrier |> mutate(
  avg_transit_days =
    0.7 + # our base handling time 0.7 days
    lane_distance_km / 850 + # driving time - 850 km per day
    if_else(lane_type == "Interstate", 0.5, 0) + # Interstate deliveries take 0.5 days longer
    rnorm(n(), mean = 0, sd = 0.4), # We add some random noise such as congestion, missed cutoffs, weather delays
  avg_transit_days = round(pmin(pmax(avg_transit_days, 0.5), 10),1)) # Clamps the transit days between 0.5 and 10

carrier |> slice_sample(prop = 1) |> head(3)
```
































