---
title: "Basic Statistics"
output: html_document
---

# 1. Basic Statistics

## 1.1 Central tendency

### 1.1.1 Mean

```{r}
# Some data
heights <- c(166, 177, 164, 167, 177)
# Average value of the vector
mean(heights) 
```

### 1.1.2 Median

```{r}
# Middle value of the vector
median(heights)
```

## 1.2 Spread of the data

### 1.2.1 Range

```{r}
# Minimum value
min(heights)
# Maximum value
max(heights)
# Second last
heights[length(heights) -1]
# Third last
heights[length(heights) -2]
```

### 1.2.2 Variance

Population variance:

$$\sigma^2 = \frac{1}{N} \sum_{i=1}^{N} (x_i - \mu)^2$$

Sample variance:

$$s^2 = \frac{1}{n-1} \sum_{i=1}^{n} (x_i - \bar{x})^2$$

```{r}
var(heights)
```

### 1.2.3 Standard deviation

Population standard deviation:

$$\sigma = \sqrt{\sigma^2} = \sqrt{\frac{1}{N} \sum_{i=1}^{N} (x_i - \mu)^2}$$

Ssample standard deviation:

$$s = \sqrt{s^2} = \sqrt{\frac{1}{n-1} \sum_{i=1}^{n} (x_i - \bar{x})^2}$$

```{r}
sd(heights)
# Alternative: sqrt(var(heights))
```

### 1.2.4 Quantile

```{r}
quantile(heights, 0.25)
```


# 2. Visualisations for Descriptive Statistics
## 2.1 Visualisations of qualitative variables
- Bar chart
- Clustered bar chart

## 2.2 Visualisations of quantitative variables
- Histogram
- Scatterplot






























