---
title: "Data Visualisation"
output: html_document
---
```{r, echo=FALSE}
# Libraries
library(ggplot2)
library(dplyr)
library(here)
library(readr)
# Get the data
Cars <- read_csv(here("ZDatasets","Cars.csv"))
```

# 1. Layered Grammar of Graphics

Quick note: `qplot()` is deprecated in order to encourage the users to learn `ggplot()` as it makes it *easier to create complex* graphics.

## 1.1 Qplot - Some Basics

Prepare the data - all this does is it relabels the variables from 0,1 to "No", "Yes".

```{r}
# Prepare the data
Cars$Sports <- Cars$Sports %>% factor(levels=c(0,1),
                                      labels=c('No','Yes'), ordered=TRUE)
Cars$Sport_utility <- Cars$Sport_utility %>% factor(levels=c(0,1),
                                                    labels=c('No','Yes'), ordered=TRUE)
Cars$Wagon <- Cars$Wagon %>% factor(levels=c(0,1),
                                    labels=c('No','Yes'), ordered=TRUE)
Cars$Minivan <- Cars$Minivan %>% factor(levels=c(0,1),
                                        labels=c('No','Yes'), ordered=TRUE)
Cars$Pickup <- Cars$Pickup %>% factor(levels=c(0,1),
                                      labels=c('No','Yes'), ordered=TRUE)
Cars$All_wheel_drive <- Cars$All_wheel_drive %>% factor(levels=c(0,1),
                                                        labels=c('No','Yes'), ordered=TRUE)
Cars$Rear_wheel_drive <- Cars$Rear_wheel_drive %>% factor(levels=c(0,1),
                                                          labels=c('No','Yes'), ordered=TRUE)
Cars$Cylinders <- Cars$Cylinders %>% as.factor()
```

### 1.1.1 Quick `qplot()` plots

```{r}
qplot(x = Cylinders, data = Cars, geom ="bar")
```

### 1.1.2 Make a box plot

```{r}
qplot(x = Cylinders, y = Kilowatts, data = Cars, geom = "boxplot")
```

### 1.1.3 Make a scatter plot

```{r}
qplot(x = Weight, y = Economy_city, data = Cars, geom = "point")
```

### 1.1.4 Transform variables

```{r}
qplot(x = log(Weight), y = log(Economy_city), data = Cars, geom = "point")
```

### 1.1.5 Map other aesthetics

```{r}
qplot(x = log(Weight), y = log(Economy_city), data = Cars, geom = "point", colour = Cylinders, shape = Cylinders)
```

### 1.1.6 Trend lines

```{r}
qplot(x = log(Weight), y = log(Economy_city), data = Cars, geom = "point") +
  stat_smooth(method = "lm")
```

### 1.1.7 Facet

```{r}
# First group categories
Cars_filter <- Cars |> filter(Cylinders %in% c("4", "6", "8"))
qplot(x = Kilowatts, y = Retail_price, data = Cars_filter, geom = "point", facets = Cylinders ~ .) +
  stat_smooth()
```

- `Cylinders ~ .` –> the left-hand side of the tilde symbol refers to rows and the right-hand side to columns.
- `.` signifies that no facets will appear on that dimension. In other words, split `Cylinders` across rows.
- If we wanted to move to columns, we would use `. ~ Cylinders` or just `~Cylinders`.

## 1.2 GGPlot - Main Usage

1. Data - the dataset being visualised
2. Aesthetics (aes) - mapping data variables to visual properties
3. Geometries (geom) - the type of plot element - e.g. points, lines, bars
4. Statistics (stats) - summarise data
5. Coordinates (coord) - coordinate system, e.g. Catesian, polar
6. Facets - split data into subplots by variable
7. Scales - control mapping from data to aesthetics (color, axes, transformations)
8. Theme - control non-data elements (fonts, background, gridlines)

```{r}
# Example - ggplot2 already loaded
ggplot(data = mtcars, aes(x = wt, y = mpg, color = factor(cyl))) +  # data + aesthetics
  geom_point(size = 3) +                                            # geometry layer
  geom_smooth(method = "lm") +                                      # statistical layer
  labs(title = "MPG vs Weight by Cylinder") +                        # labels
  theme_minimal()                                                    # theme layer
```


?qplot













