---
title: "Generating synthetic data"
output: html_document
---

```{r}
# setting a seed
set.seed(123)
# Libraries
library(magrittr)
library(dplyr)
```


# 1. Generate random data from a given interval - Uniform Distribution

## 1.1 Generating continuous uniform data

```{r}
# Generate 8 continuous values between 0 and 100
rand1 <- runif(8, 0, 100)
rand1
```

### 1.1.1 Adding syntethic data to a DataFrame

```{r}
# Create a df
df <- data.frame(customer_id = 1:5,
                 customer_name = c("Ted", "Barry", "Eli", "Adam", "Jane"))
# Adding a column
df |> 
  mutate(sale_amount = runif(5, 10, 50) |> round(digits = 2))
```

### 1.1.2 Visualising the uniform distribution

```{r}
rand2 <- runif(100000, 0, 100)
hist(rand2)
```


## 1.2 Generate integer-valued random data - discrete uniform distribution

```{r}
# Can be used for numerical and categorical values
some_categories <- c("red", "blue", "green", "orange", "purple")
rand3 <- sample(0:10, 5, replace=FALSE)
rand4 <- sample(some_categories, length(some_categories), replace=TRUE)
rand5 <- sample(c("head","tail"), 5, replace = TRUE)
rand3
rand4
rand5 # coin flip
```

# 2. Generate random data - Normal distribution

```{r}
nom1 <- rnorm(n = 200, mean = 5.0, sd = 2.5)
nom1[1:4]
```

```{r}
hist(nom1, breaks = 30)
```

## 2.1 Calculate the z-score

```{r}
z_scores <- nom1 |> outliers::scores(type = 'z')
hist(z_scores, freq = FALSE, breaks = 30)
```

## 2.2 Q-Q plot

Should appear approximately a straight line if the data are Normal.

```{r}
# QQ plot
qqnorm(nom1, main = "Q-Q Plot of Synthetic Data")
qqline(nom1, col = "red", lwd = 1, lty = 2)
```

Majority of the data lies on the QQ-plot expect for its tails.

## 2.3 Shapiro Test

Tests the Null Hypothesis that the data are distributed normally. A significant p-value means we must reject the Null Hypothesis and therefore assume that the data are not Normal and vice versa.

	•	If p-value < 0.05 → Reject H₀ → Data not normal
	•	If p-value ≥ 0.05 → Fail to reject H₀ → Data normal

```{r}
# Shapiro Wilk Test from the stats package
shapiro.test(nom1)
```

Is our significance level 5% (0.05), we would fail to reject the Null Hypothesis and our data is normally distributed.









