---
title: "Sampling"
output: html_document
---

```{r, message=FALSE}
library(here)
library(tidyverse)
```


# 1. Sampling in R

- Sampling distribution: Distribution of all the possible values of a sample statistics for a given sample size selected from a population

## 1.1 Sampling distribution of the means
- Distribution of means $\hat{x}$ created by taking many repeated random samples of size n. 
- The mean of the sampling distribution is the population mean  $\mu_{\hat{x}}$. 
- The standard deviation of this distribution of the sample means is known as the standard error (SE) or the SE of the mean.

$$SE_{\text{of the mean}} = \sigma_{\hat{x}} = \frac{\sigma}{\sqrt{n}}$$
-> SE decreases as the sample size increases.


# 2. Central Limit Theorem

CLT implies that a population mean can be estimated by obtaining a large enough representative sample from the larger population. 

- When sampling from a population with mean $\mu$ and standard deviation $\sigma$, the sampling distribution of the sample mean will tend to a normal distribution with mean $\mu$ and standard deviation $\sigma / \sqrt{n}$ as the sample size becomes large ($n > 30$).


# 3. Sample size

Calculation:

$$n = (\frac{Z_{\alpha/2}*\sigma}{E})^2$$
Where:
- $E$ = Margin of error that we are willing to accept
- $1 - \alpha$ = the probability that the magnitude of that error will not exceed

```{r}
# Create the function to determine the sample size
samp_size <- function(z, E, sigma){
  out <- ((z * sigma) / E)^2
  out
}
```

```{r}
# Create our variables
alpha <- 0.05 # for a 95% CI, alpha is 5%
sigma <- 0.03 # the standard deviation of the chain of shops
E <- 0.01 # The expected effect size (MoE)
z <- qnorm(alpha/2, lower.tail = FALSE)

# Example
n <- samp_size(z = z, E = E, sigma = sigma)
print(ceiling(n)) # round up to the next integer
```

# 4. Confidence intervals
## 4.1 Confidence intervals

Allow you to gauge the level of confidence you can attach to a point estimate like mean or proportion.

- Is the population standard deviation known, use z-score
- Is the population standard deviation unknown, use t-score and substitude with the sample standard deviation $s$. Also $n-1$.

## 4.1.1 Confidence interval for a population mean

$$\text{CI (SD known) = }\hat{x} \pm z_{\alpha/2}\frac{\sigma}{\sqrt{n}}$$
- Standard error: $\frac{\sigma}{\sqrt{n}}$
- Z = $\frac{\hat{X}-\mu}{\sigma / \sqrt{n}}$

### Example Z-Score

```{r}
# Example
alpha = 0.05 # 1 - alpha of 95%
Z = 1.96 # Z-score
x_hat = 172.05 # Point estimate xhat
sd = 6.35 # Standard deviation
n = 10 # Sample
# Calculation
margin_of_error = Z * (sd/sqrt(n))
lower_limit = x_hat - margin_of_error
upper_limit = x_hat + margin_of_error
glue::glue("A 95% confidence interval (known SD) 
           limit goes from {round(lower_limit,2)}cm to {round(upper_limit,2)}cm for the self-reported height of men.")
```

### Example t-score

$$\text{CI (SD unknwown) = }\hat{x} \pm t_{n-1,\alpha/2}*\frac{s}{n-1}$$

We use the sample metrics as above:
- X_hat of 172.05
- n = 10
- Significance level of 95% (1 - alpha / 2 in each tail)
- BUT: Sample standard deviation now: 6.06

```{r}
# Calcualte the t-distribution and new margin of error
sd_sample = 6.06
t_score = qt(0.05/2, 9, lower.tail=FALSE) # alpha / 2 for each tail , n - 1 = 9
margin_of_error_new = t_score * (sd_sample/sqrt(n))
lower_limit_new = x_hat - margin_of_error_new
upper_limit_new = x_hat + margin_of_error_new
# Result
glue::glue("A 95% confidence interval (unknwon SD) 
           limit goes from {round(lower_limit_new,2)}cm to {round(upper_limit_new,2)}cm for the self-reported height of men.")
```

## 4.2 Confidence interval for a Population Proportion
Examples:
- What proportion of Australians are willing to take out a mortgage for up to five times their annual income?
- What percentage of children aged 8-11 have their own smartphone?
- What percentage of the population owened a car before they were 20?
- Standard error of the point estimate for p = $\hat{p} = \sqrt{\frac{\hat{p}(1-\hat{p})}{n}}$
- Sample proption $\hat{p} = \frac{\text{positive responses}}{\text{all responses}}$ 

$$\text{CI_proportion} = \hat{p} \pm Z * \sqrt{\frac{\hat{p}*(1-\hat{p})}{n}}$$

```{r}
# Example
# Imagine we are interested in the proportion of adults in Melbourne who watch movies on a tablet. A random survey was conducted which found that of the 400 people surveyed, 342 responded yes – they watch movies on a tablet. Compute a confidence interval estimate using a 95 per cent confidence level for the true proportion of adults who watch movies on a tablet. They either watch movies on a tablet or they don’t, so we know we are dealing with a binomial distribution where n = 400. Here then are the steps in calculating the confidence interval.

# Values
n <- 400
phat <- 342 / n # Sample proportion
SE_phat <- sqrt((phat * (1 - phat)) / n) # Standard Error
Z <- qnorm(0.05/2, 0, 1, lower.tail=FALSE) # Z-Score
# CI propotion
lower_limit_pro <- phat - Z * SE_phat
upper_limit_pro <- phat + Z * SE_phat
glue::glue("We estimate with 95% confidence that the true proportion of adults in Melbourne who watch movies on their tablets is between {round(lower_limit_pro,2)}% and {round(upper_limit_pro,2)}%.")
```

### Sample size of proportions

$$n = (\frac{Z_{\alpha/2}^2*\sigma^2}{ME^2})$$
where:
$\sigma^2 = p(1-p)$

```{r}
# Example for sample means
# Suppose you want to derive a 95% confidence interval for the average number of hours per week that 8–11-year-old kids spend watching TV, where the margin of error is around 1 hour. You conduct a small pilot study and obtain a sample standard deviation of 3.5 hours.

n = ((1.96^2)*(3.5)^2)/1^2
ceiling(n)
```

```{r}
# Example for sample proportion
# Suppose you want to derive a 95% confidence interval for the proportion of 8–11-year-old kids who are allowed to take their smartphone to bed, where the margin of error is 10%. You conduct a small pilot study and obtain an initial estimate of 40% being allowed to take their smartphone to bed.

p = 0.4
n = ((1.96^2)*p*(1-p))/0.1^2
ceiling(n)
```


# 5. Hypothesis testing
## 5.1 Fundamentals of hypothesis testing

Six Step procedure:
1. Frame your competing hypotheses in terms of parameter of interest
2. Assume the null hypothesis is true $H_0$
3. Choose your significance level $\alpha$
4. Calculate the observed value of the test statistic from your random sample
5. Calculate the p-value and compare it with your significance elvel
6. Reject your null hypothesis if and only if the p-value < significance level $\alpha$

Which hypothesis test to use? For Z-test you need:
- Sample size greater than 30
- Known population standard deviation
-> Is one not fulfilled, use t-test.

## 5.2 Hypthesis testing for one sample

- One-tail hypothesis test: Used, when interested in whether the value of the parameter of interest is less the value specified in the null hypothesis *OR* parameter of interest is greater the value specified in the null hypothesis (NOT BOTH). $H-0: \mu = 0$ versus $H_a: \mu > 0$ *OR* $H_a: \mu < 0$
- Two-tail hypothesis test: Used, when interested in whether the value of the parameter of interest differs (either above or below) from the value specified in the null hypothesis. $H_0: \mu = 1000$, $H_a: \mu \neq 1000$

### Applying a z-test for one sample

```{r}
# The aim of this example is to test the hypothesis that people working in profession 7 have an IQ higher than average, at a significance level of 5%. Note that IQ test scores are calculated based on a normed group with average 100 and a standard deviation 15. 
iq_df <- read.csv(here("ZDatasets","IQ.csv"))
head(iq_df)
```

```{r}
# Subset profession 7
profession7 <- iq_df |> filter(Profession == 7)
# Check the sample size
profession7$IQ |> length()
```

Since the population SD is known and the sample size is large, a z-test is conducted.

```{r}
population_sd <- 15
population_mu <- 100
n <- profession7$IQ |> length()
x_bar <- mean(profession7$IQ)
SE <- population_sd / sqrt(n)
z_score <- (x_bar - population_mu) / SE
sig_level <- 0.05

# Upper tail test has p-value = P(Z > z)
p_value = pnorm(z_score, lower.tail=FALSE)
cat(paste("p-value: ", p_value))
```

Based on the p-value and the significance level (0.05), we can conclude our findings.

```{r}
statement = c("the null hypothesis at the 5% significance level.",
"\nConclude that those working in profession 7 have")
true = c("Reject", statement, "\na higher average IQ than the overall population.")
false = c("Fail to reject",statement, "\nthe same IQ as the overall population.")
if (p_value < sig_level) cat(paste(true)) else cat(paste(false))
```

































